% main.tex, to be used with thesis.tex
% This contains the cooperation chapter


\chapter{Manipulation}

Much of the innovation in the history of IPD came from various refinements of the Nash equilibrium. Reframing the equilibrium in terms of ecology led to Axelrod's idea of collective stability and his analysis of the success of TIT FOR TAT. An evolutionary refinement of by Maynard Smith allowed the introduction of mutations and a stability robust to them. It exposed a weakness of TFT---it could be invaded by a pair of strategies. To account for possibility of error or miscommunication, Richard Selten \cite{Selten_1975} proposed the trembling hand perfect equilibrium, and simulations showed that a simple strategy WSLS performed well in this error-prone environment. Not all successful strategies have the simple elegance of TFT and WSLS, however. In 2010 Iliopoulos, Hintze, and Adami showed that stability of strategies depended on the mutation rate, and analyzed a strategy they called GENERAL COOPERATOR (GC) with $\mathbf{q}_{\textrm{GC}} = (0.935, 0.229, 0.266, 0.42)$, which was the evolutionary fixed point at low mutation rates \cite{iliopoulos_2010}.

In 2012 a new class of strategies was born out of yet another approach to solving the dilemma. If the IPD is restricted to memory-one strategies, which can only use the previous state of the game in each decision, it can be thought of as a Markov process with states CC, CD, DC, DD. Since the strategies can be described as vectors of probabilities the player will cooperate given each of the four states of the game
\begin{align*}
\mathbf{p} &= (p_\CC, p_\CD, p_\DC, p_\DD) \\
\mathbf{q} &= (q_\CC, q_\DC, q_\CD, q_\DD)
\end{align*}
where $p_2 = \mathds{P}(\textrm{player 1 cooperates } | \CD)$. Then the Markov transition probabilities are
\[
\left[
\begin{array}{c c c c}
p_1q_1 & p_1 (1-q_1) & (1-p_1)q_1 & (1-p_1)(1-q_1) \\
p_2q_3 & p_2 (1-q_3) & (1-p_2)q_3 & (1-p_2)(1-q_3) \\
p_3q_2 & p_3 (1-q_2) & (1-p_3)q_2 & (1-p_3)(1-q_2) \\
p_4q_4 & p_4 (1-q_4) & (1-p_4)q_4 & (1-p_4)(1-q_4)
\end{array}
\right]
.\]

If $\mathbf{v}$ is the stationary distribution of the above chain, then we can define the long term average payoffs to be
\begin{align*}
S_X &= \mathbf{v} \cdot \mathbf{S}_X = \mathbf{v} \cdot (\sR, \sS, \sT, \sP) \\
S_Y &= \mathbf{v} \cdot \mathbf{S}_Y = \mathbf{v} \cdot (\sR, \sT, \sS, \sP).
\end{align*}
Press and Dyson's discovery was that the dot product of the stationary distribution $\mathbf{v}$ with an arbitrary 4-vector $\mathbf{f}$ is given by the following determinant
\[
\mathbf{v}\cdot \mathbf{f} \equiv D(\mathbf{p}, \mathbf{q}, \mathbf{f})
= \det
\left[
\begin{array}{c c c c}
-1 + p_1 q_1    &	-1 + p_1	&	-1 + q_1	&	f_1 \\
p_2 q_3         &	-1 + p_2	&	-q_3		&	f_2 \\
p_3 q_2 		&	p_3			&	-1 + q_2	&	f_3 \\
p_4 q_4			&	p_4			&	q_4			&	f_4
\end{array}
\right],
\]
in which the second column is $\tilde{\mathbf{p}} \equiv (-1 + p_1, -1 + p_2, p_3, p_4)$ is entirely controlled by the first player, and the third column $\tilde{\mathbf{q}} \equiv (-1 + q_1, -p_3, -1 + q_2, q_4)$ is entirely controlled by the second player. This allows each player to unilaterally make the determinant vanish by setting their column to be a scalar multiple o $\mathbf{f}$. The opportunity for roguery arises when we consider $\mathbf{f} = \alpha \mathbf{S}_X + \beta \mathbf{S}_Y + \gamma\mathbf{1}$. Since $\mathbf{v} \cdot  (\alpha \mathbf{S}_X + \beta \mathbf{S}_Y + \gamma\mathbf{1}) = \alpha S_X + \beta S_Y + \gamma$, each player can unilaterally enforce a linear relation between long the players' long term average payoffs by setting determinant to zero. The strategies that enforce this linear relationship were christened \textit{zero determinant (ZD)} strategies.

Press and Dyson focused on two types of ZD strategies. The first allows X  to unilaterally set her opponent's score by setting $\alpha = 0$ and thus forcing $\beta S_Y + \gamma = 0$. Solving the equations
\begin{align*}
-1 + p_1 &= \beta \sR + \gamma \\
-1 + p_2 &= \beta \sT + \gamma \\
p_3 &= \beta \sS + \gamma \\
p_4 &= \beta \sP + \gamma
\end{align*}
to eliminate the parameters $\beta$ and $\gamma$ and expressing $p_2$ and $p_3$ as functions of $p_1$ and $p_4$, they arrive at the strategy
\begin{align*}
p_2 = \frac{p_1(T - P) - (1+p_4)(T-R)}{R-P} \\
p_3 = \frac{(1-p_1)(P-S)+p_4(R-S)}{R-P}
\end{align*}
which sets Y's score to a weighted average of $P$ and $R$
\[
S_Y = \frac{(1-p_1)P + p_4 R}{(1-p_1) + p_4}.
\]
This equation has feasible solutions when $p_1$ is close to but not equal to 1, and $p_4$ is close to but not equal 0.

Using a similar calculation Press and Dyson show that $X$ cannot set her own score with $\tilde{\mathbf{p}} = \alpha \mathbf{S}_X + \gamma \mathbf{1}$, as
\[
p_2 = \frac{(1+p_4)(R-S) - p_1(P-S)}{R-P} \geq 1
\]
except for $\mathbf{p} = (1, 1, 0, 0)$.


The second type of ZD strategies described by Press and Dyson is even more devilish. It allows X to demand and get an extortionate share of the payoff surplus over the mutual defection value $P$ by setting
\[
\tilde{\mathbf{p}} = \phi [(\mathbf{S}_X - P\mathbf{1}) - \chi (\mathbf{S}_Y - P \mathbf{1})]
\]
which leads to
\begin{align*}
p_1 &= 1 - \phi(\chi-1) \frac{R-P}{P-S} \\
p_2 &= 1 - \phi\left(1 + \chi \frac{T-P}{P-S}\right) \\
p_3 &= \phi (\chi + \frac{T-P}{P-S}) \\
p_4 &= 0
\end{align*}
with feasible strategies existing for any $\chi$ and
\[
0 < \phi \leq \frac{P-S}{(P-S) + \chi(T-P)}.
\].
Computing the long-run average scores for X and Y with using Axelrod's values $(T, R, P, S) = (5, 3, 1, 0)$ Press and Dyson show
\[
S_X = \frac{2 + 13\chi}{2 + 3\chi} > 3 = P\textrm{, and } S_Y = \frac{12 + 3\chi}{2 + 3\chi} < 3 = P
\]
for $\chi > 1$, while the limiting fair case $\chi = 1$ and $\phi = 1/5$ reduces to TIT FOR TAT with $\mathbf{p}^\textrm{TFT} = (1, 0, 1, 0)$.

When playing against an extortioner X, Y has two choices, the first is to try to maximize his own score, but that would feed into X's plot and increase X's score even more. The second is to refuse to be extorted by playing AllD either out of spite or in hopes that X recognizes that extortion will not work and switches to a more equitable play. The key to understanding a match between two extortioners is in the $p_4 = q_4 = 0$. Once they reach a mutual defection, they will stay there indefinitely thus receiving a long term average of $P$. This should raise questions about evolutionary stability of extortionate ZD strategies. In the two years since Press and Dyson published their findings, a number of papers came out discussing how well can ZD strategies fare in ecological and evolutionary settings. Before moving on to those results, however, I should note two more theorems established by Press and Dyson.

The first shows that restricting our attention to memory-one strategies is not as limiting as it may seem. They prove that for any finite memory strategy there exists a memory-one strategy... \cmt{review}. In their 2012? paper Adami and Hintze \cite{Adami_Hintze_2014} note that longer memory may provide evolutionary advantage to variants of ZD players who attempt to recognize each other by analyzing longer pattern of play and switch to mutual cooperation in order to increase their evolutionary fitness.

The second ancillary result dispels doubts that may arise about the steady state payoff $S_i = \mathbf{v}\cdot \mathbf{S_i}$, $i \in \{X, Y\}$. It may take hundreds of moves for the linear relationship
\[
\alpha S_X + \beta S_Y + \gamma = 0
\]
to be established, and one may wonder if there is any way the non-ZD player can prevent that relationship by changing their own play inside the equilibration time scale \cmt{find better wording}. Press and Dyson answer that question negatively, thus showing that in a one-on-one play the non-ZD player cannot benefit from having longer memory or by keeping the game away from the Markov stationary distribution and set the course for the future research to be focused on memory-one strategies, which significantly simplifies the space of strategies and the search for its structure.

\subsubsection{What can ZDs do?}

Since ZD strategies may take hundreds of turns for the average payoffs to settle near their expected values, they may not be particularly useful in modeling situations where interactions are not frequent or the expected number of interactions is low, like human behavioral experiments or duopoly output. It is easier to imagine these behaviors to arise in evolutionary or ecologic systems where numbers of interactions over thousands of generations may be much higher. One needs not assume that players in evolutionary systems are aware of ZD strategies and are capable of consciously calculating the probabilities $(p_1, p_2, p_3, p_4)$ if ZD strategy could evolve as a result of natural selection. The possibility of evolutionary dynamics giving birth to ZD strategies and their stability once they are introduced to a population is the subject of a number of research papers published shortly after Press and Dyson made their discovery. The remainder of this paper presents some of these results.

Within a few month of Press and Dyson's publication, Stewart and Plotkin \cite{Stewart26062012} performed a simple experiment. They recreated Axelrod's first round-robin tournament with several additional players: GENEROUS TIT FOR TAT (GTFT) (recall that GTFT would have won Axelrod's first tournament had it been submitted \cmt{check that I mentioned that in Axelrod section}), the compliant \textit{zero determinant generous tit for tat 2 (ZDGTFT-2)} which forced the relationship $S_X - R = 2(S_Y - R)$ so that $S_X \leq S_Y$, and \textit{Extort-2} with $\mathbf{p}^{\textrm{E2}} = (8/9, 1/2, 1/3, 0)$. As one may expect, Extort-2 won the highest number of matches save for AllD. Those victories, however, were mostly pyrrhic. Extort-2's average score was second to last (again bested, or rather worsted, only by AllD). More surprisingly ZDGTFT-2 received the highest average score followed by Axelrod's usual suspects GTFT, TFT, and TF2T. 

Stewart and Plotkin's simulations suggest that extortioners do not do well when faced with a population consisting of a variety of strategies. This hypothesis was further explored by Christoph Adami and Arend Hintze in a paper that first appeared in August 2012 and has since undergone several revisions \cite{Adami_Hintze_2014}. They analyzed evolutionary performance of the two types of ZD strategies discussed by Press and Dyson, the dictatorial \cmt{they are called equalizers} $ZD^D$ that unilaterally sets the opponent's score, and the extortioner $ZD^E$. Except for the limiting case with $p_1 = 1$, the dictatorial $ZD^D$ strategies do not fare well against their own copies, as they force on their clones the same score $E(ZD^D | ZD^D) \leq R$ that they force on all other strategies, and thus are evolutionarily dominated by TFT, Pavlov, and other strategies that receive a score of $R$ against their own kind. Extortioners suffer from even poorer performance against other extortioners, since $p_4 = 0$ inexorably drives them to mutual defection and the long term average of $P$. In two-strategy competitions some strategies do give $ZD^D$ the upper hand. In particular the general cooperator GC which Iliopoulos, Adami, and Hintze showed to be a fixed point at low mutation rates \cite{iliopoulos_2010} is dominated by ZD with $E(Z, GC) = 2.125$ and $E(GC, GC) = 2.11$. However this advantage does not contradict their earlier findings. Agent based simulations with mutation rate favoring GC and seeded with the dictatorial ZD ($p_1 = .99$, and $p_4 = .01$) showed that the $ZD^D$ strategy evolves into GC \cite{Adami_Hintze_2014}, and is thus evolutionarily and mutationally unstable. In closing, Adami and Hintze concede that extortionate ZDs may be evolutionarily fit if they learn to recognize each other either through a tagging mechanism or by having access to longer play history (recall that ZDs are all memory-one strategies), and adapt to cooperate with other ZDs and extort from non-ZD strategies. But they note that other strategies could take advantage of that recognition by developing fake tags or patterns of play that would make them appear ZD and thus provoke cooperation. Nature provides countless vibrant examples of such adaptation including the familiar syrphidae flies that mimic the coloring of bees or wasps and thus protect themselves against predators.

In a 2013 paper Christian Hilbe, Martin A. Nowak, and Karl Sigmund analyzed a special case of prisoner's dilemma called the donation game with the payoff structure given by
\[
\left(
\begin{array}{c c}
b-c & -c\\
b & 0
\end{array}
\right)
\]
and the pool of strategies consisting of $TFT$, $E_\chi$, $WSLS$, $AllC$, and $AllD$. They note that in pairwise comparisons $E_\chi$ and $AllD$ are neutral, $TFT$ weakly dominates $E_\chi$ by playing better against other $TFT$ than $E_\chi$ plays against other extortioners, and that $AllC$ and extortioners can invade each other and stably coexist in proportion $c(\chi-1):(b+c)$, and finally $WSLS$ dominates $E_\chi$. $TFT$ can always invade mixed equilibria of extortioners and unconditional cooperators or defectors, but can in turn be invaded by other nice strategies.

To study equilibrium distributions of these strategies in finite populations, Hilbe et al. modeled natural selection as an imitation process. At each step two randomly chosen players $X$ and $Y$ compare their average payoffs $S_X$ and $S_Y$, and $Y$ switches to $X$'s strategy with some probability that is a function of the difference $S_X - S_Y$. Although in this model extortioners were never the most abundant outcome, they played an interesting role of catalyzing cooperation. Populations made up solely of $WSLS$ and $AllD$ were stuck in stable equilibria containing predominantly $AllD$, however when $E_\chi$ or $TFT$ were added to the mix, equilibrium distributions favored $WSLS$ in all but the smallest populations. The results were similar with the limiting case of rare mutations (with equilibria computed analytically using the method of Fudenberg and Imhof \cite{Fudenberg_2006}) and more frequent mutations (equilibria computed by agent-based simulations). Except in small populations, rare mutations led to $WSLS$ fully overtaking the population and driving $AllD$, $E_\chi$, $AllC$, and $TFT$ extinct. When mutations were more frequent, the populations stabilized at mixtures heavily favoring $WSLS$. In these scenarios extortionate ZDs (including the fair TFT) were shown to be catalysts of cooperation. They interestingly complement and deepen the previous results discussed here, including Axelrod's chronology of emergence of cooperation \cite[p.55]{axelrod1984evolution} and long-term dominance of WSLS discovered by Nowak and Sigmund \cite{Nowak1993}. It is also a partial rebuttal to Adami and Hintze's assertion that `winning isn't everything' \cite{Adami_Hintze_2014}, as Hilbe et al. show that evolutionarily unstable strategies can nonetheless play a vital role in evolution of cooperation.

Hilbe et al. present another scenario in which extortioners prosper. When the IPD is played between two populations (hosts and their symbionts), or two types within the same populations (buys and sellers, males and females, constituents and representatives), extortion strategies can evolve even in large populations because their low scores when playing their own type does not hinder their fitness. If the first group plays $E_\chi$ and the second group attempts to maximize their own score by playing the best response, the second group will adopt $AllC$. However in this situation the first group is likely to evolve an even more profitable $AllD$ which would force the second group to also switch to $AllD$. An interesting special case was revealed by simulations in which the two populations evolved at different rates of mutations. In a simulation seeded with non-ZD strategies the slowly evolving host populations arrived to a $\delta = 0.1$ neighborhood of $E_\chi$ (Euclidean distance on the 4-cube $(p_1, p_2, p_3, p_4)$), and were able to extract a surplus more than 10-fold larger than what was achieved by the rapidly evolving symbionts. The hosts' payoff after 2000 generations was consistently above the mutual cooperation value $R = b-c = 2$, while the symbionts received scores close to $P = 0$. \cmt{note host and symbiont higher in paragraph}

The discovery of ZD strategies by Press and Dyson motivated a number of papers that focused on extortionate behavior, however it could be that the more successful ZDs are of an entirely different type -- the compliers, whose strategy is defined \cmt{reference equation from previous Stewart and Plotkin}. In their previous paper, Stewart and Plotkin showed that a compliant strategy they called ZDGTFT-2 won an Axelrod-style tournament that contained TFT, GTFT, TF2T, WSLS and a number of other strategies. Using the criterion of \textit{evolutionary robustness}, a modification of ESS to finite populations, they explored the subset of generous ZD strategies (of which ZDGTFT is an example) and found that generous ZDs perform very well in simulations, and in some conditions even outperform WSLS \cite{Stewart03092013}. The chronology of research following Press and Dyson's introduction of ZD strategies is reminiscent of Axelrod's tournaments. Following the success of TFT the players were looking to gain an upper hand by submitting variants of TFT that attempted to take advantage of opponents by not always being nice. These versions almost invariably did worse than TIT FOR TAT, but also worse than more generous variants TF2T and GTFT. Similarly the initial interest in ZDs focused on extortionate and dictatorial equalizer strategies designed to gain at cost to opponent, when in reality ZD strategies that practiced generosity, the opposite of extortion, turned out to give their practitioners better evolutionary outcomes. I find that these results offer an optimistic view of nature, and a somewhat pessimistic view of humanity as represented by the community of evolutionary game theorists whose first instinct, it seems, is to opportunistically defect.

\chapter{Other results and further reading}
In preparation of this exposition I found some results that due to considerations of time and volume I did not include in this paper. In 1990s Jean Pierre Langlois, who advised me on this paper, introduced a decomposition theorem that showed that subgame perfect equilibria... he called these strategies \textit{countervailing}. One of the properties of countervailing strategies is that they, similarly to equalizer ZDs, can unilaterally set their opponent's score. This led us to wander about the nature of the relationship between the space of countervailing strategies and the space of ZDs, and we found an example of an intersection.

In 2013 several papers explored the structure of the space of memory-one strategies and ZD strategies. Some of them are cited above, others are ...

Finally this summer Tyler Singer-Clark under the supervision of Scott Aaronson published his senior thesis in which he used the setting of an IPD tournament and an algorithm similar to PageRank to analyze how moral various strategies are if we assume that moral players cooperate with other moral players and defect against immoral players. There is an inherent circularity in this definition of morality, since the only way we classify a player is moral if he is good to other moral players, who in turn have to be good to other moral players etc. PageRank was designed to circumvent a similar circularity in the link structure of the internet, where a page was deemed important if it was linked to by other important pages, which, in turn were deemed important if they had good incoming links and so on.





\section{Results}

\subsubsection{IPD contains strategies that dominate any evolutionary opponent, Press, Dyson \cite{Press26062012}
\\(April 19, 2012)}
This paper introduced ZD strategies and talked about two types of ZDs. Extortioners (including the fair extortioner TFT), and strategies that unilaterally set opponent's score. The scenario that Press and Dyson describe is an infinite play between two opponents. If one of the opponents knows about ZD and the other attempts to optimize his score, then the ZD opponent can extort. However if both opponents play ZD, there is no surplus to share (this is easy to see since $p_4 = q_4 = 0$ for extortioners). 


\subsubsection{Stewart and Plotkin \cite{Stewart26062012}
\\(June 26, 2012)}
Stewart and Plotkin plug in two ZD strategies in Axelrod's first tournament in a short follow-up paper \cite{Stewart26062012}. The strategies are Extort-2 with $S_X - P = 2(S_Y - P)$ that guarantees player X twice the share of payoffs above P compared with those received by Y. ZDGTFT-2 forces $S_X - R = 2(S_Y - R)$ offers Y a higher portion. ZDGTFT-2 won by average score. Extort-2 won the highest number of matches (except ALLD).

\subsubsection{Evolutionary instability of Zero Determinant strategies demonstrates that winning isn't everything, Adami, Hintze \cite{Adami_Hintze_2014}
\\(August 13, 2012 (v1)) (published in Nature Communications in 2013)}

Adami and Hintze study evolutionary stability of two types of ZD strategies defined by Press and Dyson. The first type sets the opponent's score unilaterally by following 
\[
\mathbf{p} = \left(p_1, \frac{p_1(T-P) - (1+p_4)(T-R)}{R-P}, \frac{(1-p_1)(P-S) - p_4(R-S)}{R-P}, p_4\right)
\]
with $p_1$ close to but not equal 1, $p_2$ close to but not equal 0.

They define functions $f, h, g$ to be the expected scores in a matches of ZD and some other strategy O against each other and themselves,
\begin{align*}
f(\mathbf{p})&=\mathbb{E}(ZD, ZD) = \mathbb{E}(O, ZD)\\
h(\mathbf{q}) &= \mathbb{E}(O, O)\\
g(\mathbf{p}, \mathbf{q}) &= \mathbb{E}(ZD, O)
\end{align*}
and define a new game in which each player choses whether to play ZD or O with the matrix
\[
\left(
\begin{array}{c c}
f(\mathbf{p}) & g(\mathbf{p}, \mathbf{q}) \\
f(\mathbf{p}) & h(\mathbf{q})
\end{array}
\right)
.\]
Can subtract a constant from each column \cmt{find reference}
\[
\left(
\begin{array}{c c}
0 & g(\mathbf{p}, \mathbf{q}) - h(\mathbf{q}) \\
0 & 0
\end{array}
\right)
\]
so if $g(\mathbf{p}, \mathbf{q}) - h(\mathbf{q}) > 0$ for every O, ZD is weak ESS \cmt{What is Weak ESS?}. A mixture of ZD and O cannot be mixed ESS because ESS enforces the same score on itself as it does on others.

They go on to compare ZD to PAVLOV and show that PAVLOV is the ESS \cmt{Didn't we decide that no pure strategy is ESS in Part 1? Could they be only looking at ESS given no mutations and a fixed starting finite set of strategies?} They do this using the replicator equations
\[
\dot{\pi}_i = \pi_i(w_i - \bar{w})
\]
where $\pi_i$ is the proportion of population using strategy $i$, $w_i$ is the fitness of strategy $i$, and $\bar{w}$ is the average fitness in the population. They show that independently of initial distribution of ZD and PAVLOV, over time PAVLOV takes over and ZD goes extinct. Agent-based simulations follow qualitatively identical trajectories (time scale may be different).

Another illustrative example is the performance of ZD against the ``general cooperator (GC)'' $\mathbf{q}_{GC} = (0.935, 0.229, 0.266, 0.42)$, which Iliopoulos, Adami, and Hintze showed to be an evolutionary fixed point at low mutation rates \cite{iliopoulos_2010}. $E(Z, GC) = 2.125$ and $E(GC, GC) = 2.11)$, so ZD is a weak ESS. How, then the GC is the evolutionary fixed point and not ZD? Adami and Hintze showed that if a mutation rate favoring GC is used in an agent based simulation with the ZD strategy, ZD evolves into GC. Thus ZD is not genetically or mutationally stable in addition to not being ESS.

Extortionate ZD strategies are not ESS due to their poor performance against other extortioners $p_4 = 0$ means that the game gets stuck in DD state following the first mutual defection. $ZD_E$ strategies that can recognize other $ZD_E$s can be ESS by extorting others but cooperating among themselves. Of course the recognition mechanisms can lead to their opponents evolving methods to behave like $ZD_E$s and thus subvert their advantage. \cmt{Syrphidae flies are colored like wasps or bees}.

This possibility of self recognition suggests that short-memory players cannot set the rules of the game in an evolutionary setting. Press and Dyson's claim that a long-but-finite memory strategy is equivalent to some memory-one strategy is correct, but in evolutionary setting when strategies can use memory to recognize type of opponent, longer memory can be useful, and can lead to better stability if extortioners recognize each other and cooperate.

Other scenarios of useful or successful extortion are given in the following paper (Evolution of Extortion in IPD Games).

\subsubsection{Evolution of extortion in IPD games, Hilbe, Nowak, Sigmund \cite{Hilbe23042013}
\\(April 23, 2013)}
Examine the contest between extortioners and four of the most important memory-one strategies. Show that extortion cannot be an outcome of evolution, but can catalyze the emergence of cooperation. Extortion strategies can only get a foothold if the population is very small. If IPD is played between members of two distinct populations, ZD strategies can emerge in the population that evolves more slowly. In particular, extortion strategies can allow host species to enslave their endosymbionts.

WSLS dominates extortioners. 


\subsubsection{From extortion to generosity, evolution in the IPD, Stewart, Plotkin \cite{Stewart03092013}
\\(July 25, 2013)}

Extortion makes for a good headline, but there are ZD strategies that promote cooperation \cmt{Can use this as a way to bring the discussion back to Axelrod's arguments for TFT}

The aftermath of discovery of ZD strategies was a volume of research into 
extortioners, while generous ZDs did not attract as much attention. It seems people always are trying to find clever ways to undermine their opponents in order to beat the game, these ways did not get them very far in Axelrod's tournaments, and once again not very far in using ZDs. Generous ZDs outperformed PAVLOV and TFT, while extortionate ZDs did not do well in long run.

This paper defines another version of evolutionary robustness for finite populations.


\subsubsection{Adaptive Dynamics of Extortion and Compliance, Hilbe, Nowak, Traulsen \cite{Hilbe_extortion_compliance}
\\(November 1, 2013)}


\subsubsection{Notes on plan for Part 2}
The second body of work that I planned to use focuses on strategies that attempt to manipulate their opponents. The tentative plan for this part is to start with countervailing strategies, then go on to zero determinant strategies, explore what is the relationship between zero determinant and countervailing strategies, and then note that although zero determinant strategies are not ESS, they can perform interesting roles as catalysts of equilibrium shifts (like TFT was in Nowak and Sigmund's simulation). TFT and PAVLOV are both ZD. I will introduce evolutionary dynamics briefly only to be able to quote some results from papers that discuss stability of ZD strategies.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%






